---
title: "Non-pharmaceutical interventions for respiratory pandemics: challenges and a potential solution"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE,echo = FALSE,cache=TRUE) 
require(data.table)
require(ggplot2)
require(cowplot)
require(metR)
require(plyr)

theme_set(theme(panel.grid.major.y = element_blank(),panel.grid.minor.y = element_blank()) + theme(legend.background = element_rect(fill = "white")) + theme_half_open() + background_grid()  + theme(text = element_text(size=20), axis.text = element_text(size=20)))
```



# Frequent testing

In a homogeneous population where fraction $\gamma$ test regularly, fraction $\sigma$ of counterfactual transmissions occur after receiving a positive test, and transmissions are reduced to fraction $\beta$ after receiving a positive test, the following equation gives the effective reproductive number.

$R_e = R_0 \cdot (1 - \gamma \sigma \beta)$

$\sigma$ depends on several factors, including: frequency of testing, delay before receiving test results, test sensitivity depending on viral load, viral load trajectory over time, and behaviour change after any symptoms. With information on transmission risk and test sensitivity over time we can estimate $\sigma$.

For any infected individual $i$, data relevant to their infection can be summarized in vector $\pi_i$. This could include values like the date of infection, date of symptom onset, and potentially peak viral load. The expected transmissions on day $t$ can be estimated based on the individual data (as in Ferretti et al [https://www.medrxiv.org/content/10.1101/2020.09.04.20188516v2]) and is written as $E[T(t)|\pi_i]$, where $T(t)$ is the number of transmissions on day $t$. The probability of a test taken on day $t$ being positive can also be estimated [Petrie et al, ...], and is described as $P(Test(t)|\pi_i)$.






```{r,cache=FALSE}
# uses results from ~/MassTesting/testDetect.jl
dtTest = fread("~/MassTesting/FracTransmissions.csv")
ggplot(dtTest, aes(x = 1/TestPeriod, y = FracAfterPositive, colour =  as.factor(TestDelay))) + geom_line() + geom_point() + scale_x_continuous(expand = c(0, 0), limits = c(0,1), breaks = c(1/30, 1/7, 1/4, 1/2, 1), labels = c("1/30", "1/7", "1/4", "1/2", "1")) + scale_y_continuous(expand = c(0, 0), limits = c(0,1)) + guides(colour=guide_legend(title="Test Delay [Days]")) + xlab("Testing Frequency [1/Days]") + ylab("Fraction Counterfactual Transmissions \n After Receiving Positive Test")+ theme(legend.position = c(0.7, 0.2)) + theme(text = element_text(size=16))
```
Figure 1: Fraction of transmissions after receiving positive test as function of frequency and delay for Covid-19, conditioned on dates of infection and symptom onset.




```{r, fig.align = 'center'}


dtSweep =  data.table(expand.grid(FracTest = seq(0.5,1, length.out=200), IsoEffect =  seq(0.5,1, length.out=200), TestPeriod = c(1,4,7), TestDelay = c(0,1,2)))

dt = merge(dtSweep, dtTest, by = c("TestDelay", "TestPeriod"))
dt[, FractionR0 := (1- FracTest*FracAfterPositive*IsoEffect)]

dt[, FreqLabel := as.factor(paste0("Frequency=1/",TestPeriod))]
dt[, DelayLabel := as.factor(paste0("Delay=",TestDelay))]

breaks = seq(20, 80, by = 30)
labels = paste0(breaks, "%")


p = ggplot(dt, aes(x = FracTest, y = IsoEffect))  + geom_contour_fill(aes(z = 100*(1 - FractionR0))) + scale_fill_viridis_c(option = "magma", name = "Percent Reduction in R", breaks = breaks, labels = labels) + facet_grid(  DelayLabel ~ FreqLabel ) + theme(strip.background = element_blank()) +scale_x_continuous(labels = prettyNum) + scale_y_continuous(labels = prettyNum) + theme(legend.text = element_text(size = 10), axis.text =element_text(size = 16)) + theme(legend.key.size = unit(0.6, "cm")) + theme(legend.position="bottom") + ylab("Isolation Effectiveness") + xlab("Fraction Testing")+ theme(strip.text = element_text(size = 14))

print(p)

```

```{r}

ausIsoEffect = 0.9939

dt = dtTest[TestPeriod %in% c(1,4,7) & TestDelay %in% c(0,1,2)]

dt[, Label := paste0("Period=", TestPeriod,", Delay=",TestDelay)]

dtRe = data.table(expand.grid(FracTest = seq(0.5,1, length.out=200), Label = dt$Label, R0 = c(1.5, 4.0, 12.0 )))

dtRe = merge(dt, dtRe, by = "Label")

#dtSweep =  data.table(expand.grid(FracTest = seq(0.5,1, length.out=200), IsoEffect =  seq(0.0,1, length.out=200), R0 = c(1.5, 4.0, 12.0 )))

#dtSweep[, FracAfterPositive := (1-1/R0)/(IsoEffect*FracTest)]
#dtSweep[IsoEffect > 1, FracAfterPositive := NA]

dtRe[, IsoEffect := (1-1/R0)/(FracAfterPositive*FracTest)]
dtRe[IsoEffect > 1, IsoEffect := NA]


#showLabels = dt[(TestPeriod == 1 & TestDelay == 0)| (TestPeriod == 1 & TestDelay == 1)|(TestPeriod == 2 & TestDelay == 1)|(TestPeriod == 3 & TestDelay == 1)|(TestPeriod == 7 & TestDelay == 2),    Label]

dtRe[, R0Label := factor(paste0("R0=", R0), levels = c("R0=1.5", "R0=4", "R0=12")) ]

p = ggplot() + geom_line(data= dtRe,aes(x = FracTest, y = IsoEffect, colour = paste0("1/",TestPeriod), linetype = as.factor(TestDelay)), linewidth = 1.2) + scale_x_continuous(limits= c(0.5,1), n.breaks = 5) + scale_y_continuous(limits= c(0.5,1), n.breaks = 5 ) + geom_hline(yintercept = ausIsoEffect, colour = "purple", linetype = "dashed", linewidth = 1.2) + ylab("Minimum Necessary \n Isolation Effectiveness") + xlab("Fraction Regularly Testing")  + guides(colour=guide_legend(title="Test Frequency"), linetype=guide_legend(title="Test Delay"))+ coord_cartesian(expand=FALSE) + facet_wrap(~R0Label) + theme(strip.background = element_blank()) #+ ggtitle("Testing and isolation needed \n to keep Re below 1 (with R0 = 2.5)")
# geom_contour_fill(data = dtSweep[!is.na(FracAfterPositive) & FracAfterPositive < Inf], aes( x = FracTest, y = IsoEffect, z = FracAfterPositive), alpha = 0.5, breaks = seq(1.0, 0.5, length.out = 15)) +scale_fill_viridis_g( name = "Needed Fraction", labels = prettyNum) 
#+ geom_text(aes(0.6, ausIsoEffect, label = "Effectiveness of State Quarantine in Australia", vjust = - 1))


print(p)

ggsave(p, filename = "~/MassTesting/figures/fracTransmissions.pdf")
# 1/2.5 = 1 - g*s*b
# g = (1-1/2.5)/(s*b)






```
Figure 2: Test and isolation adherence needed to keep $R_e$ below 1 for different testing frequencies and delays (assuming immuno-naive population and no other interventions used). For $R_0 = 2.5$ pathogen with incubation, transmissibility, and test sensitivity similar to Covid-19. Dashed purple line: isolation effectiveness in Australian travel quarantine facilities.



# Cost

There are two types of cost considered: fixed and variable. Fixed costs are mainly equipment capacity needed in advance of a pandemic. Variable costs include 

(From https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0277867). 

200k usd cost in equipment, can test 3000 samples per day (without pooling, and with 5 runs per day). Equipment lifetime 7-10 years 

(From https://www.thelancet.com/journals/eclinm/article/PIIS2589-5370(21)00308-4/fulltext)

\$10 USD to test each person when using 24-1 pooled saliva protocol. (including logistics, labour, pcr reagents)



```{r}

fracDt = fread("~/MassTesting/FracTransmissions.csv")



# isoCost = 10*(100 + 200) + 0.5*3000 
computeCost = function(importRate = 1e-4 , R0 = 2.0, testFraction = 0.95, isoFraction = 0.9, testPeriod = 2, testDelay = 1, poolSize = 1, incentiveCost = 10,  logisticCost = 5, pcrCost = 40, isoCost = 5000){
  transFraction = fracDt[TestPeriod == testPeriod & TestDelay == testDelay, FracAfterPositive]
  
  testFreq = 1/testPeriod
  
  
  
  
  Rt = R0*(1 - testFraction*transFraction*isoFraction)
  
  if(Rt > 1) return(1)
  
  newInfectFrac = importRate*1/(1-Rt) # fraction new infections (import and transmit per day) relative to total pop
  
  if(poolSize == 1){
    testCost = testFreq*testFraction*(incentiveCost + logisticCost + pcrCost)
  }else{
    testCost = testFreq*testFraction*(incentiveCost + logisticCost + 1/poolSize*pcrCost + poolSize*newInfectFrac*pcrCost)
  }
  
  dailyCostPerPerson = testCost + newInfectFrac*testFraction*isoCost
  gdpPerCapita = 70e3 # gdp per person in USA
  return(dailyCostPerPerson*365/gdpPerCapita)
}

```







\iffalse
# Cost vs import rate basic
```{r}
R0 = 2

dt = data.table(expand.grid(ImportRate = 10^seq(-9, -2, length.out = 200), TestPeriod  = seq(0:4), TestDelay = 1))
dt$Cost = sapply(1:nrow(dt), function(i){
  computeCost(importRate = dt[i, ImportRate], R0 = R0, testPeriod = dt[i,TestPeriod], testDelay = dt[i,TestDelay] )
} )

ggplot(dt,aes(x= ImportRate, y = Cost, colour = as.factor(TestPeriod))) + geom_line() + scale_x_log10()+scale_y_log10(limits = c(0.01, 0.5), n.breaks = 8)
```
\fi

#cost vs import rate with pooled testing and realisticly cheaper steps
```{r}
fracDt = fread("~/MassTesting/FracTransmissions.csv")
R0 = 2
incentiveCost = 5
logisticCost = 8
poolSize = 24
pcrCost = 48
isoCost = 5e3
testFraction = 0.95
isoFraction = 0.9

dt = data.table(expand.grid(DailyFracInfected = 10^seq(-7, -1, length.out = 200), TestPeriod  = seq(0:4), TestDelay = 1))
dt = rbindlist(llply(1:nrow(dt), function(i){
  
  dailyInfections = dt[i, DailyFracInfected]
  testPeriod = dt[i,TestPeriod]
  testDelay = dt[i,TestDelay]
  

  
  transFraction = fracDt[TestPeriod == testPeriod & TestDelay == testDelay, FracAfterPositive]
  
  testFreq = 1/testPeriod
  
  Re = R0*(1 - testFraction*transFraction*isoFraction)
  

  if(poolSize == 1){
    testCost = testFreq*testFraction*(incentiveCost + logisticCost + pcrCost)
  }else{
    testCost = testFreq*testFraction*(incentiveCost + logisticCost + 1/poolSize*pcrCost + poolSize*dailyInfections*testPeriod*pcrCost)
  }
    
  gdpPerCapita = 70e3 # gdp per person in USA
  
  dailyTestCostPerPerson = testCost*365/gdpPerCapita
  dailyIsoCostPerPerson = dailyInfections*testFraction*isoCost*365/gdpPerCapita
  
  dailyTotalCost = dailyTestCostPerPerson + dailyIsoCostPerPerson



  return(data.table(Cost = dailyTotalCost, TestCost = dailyTestCostPerPerson, IsoCost = dailyIsoCostPerPerson, Re = Re, TestPeriod = testPeriod, TestDelay = testDelay, FractionInfectedDaily = dailyInfections))
 
  

} ))


#dt[, Capacity := 2*1/24]
dt[ , CapacityNeeded := pmin(1/TestPeriod, 1/TestPeriod*(1/24 + 24*TestPeriod*FractionInfectedDaily))]
#dt[CapacityNeeded > Capacity, Cost := NA]
dt = dt[TestPeriod <= 3]
dt[,Description := paste(TestPeriod, "(Re=")]

dtLong = melt(dt, id.vars = c(
  "FractionInfectedDaily","TestDelay", "TestPeriod"), measure.vars = c("Cost", "TestCost", "IsoCost") )


p1 = ggplot(dtLong, aes(x= FractionInfectedDaily, y = value, colour = as.factor(TestPeriod), linetype = variable)) + geom_line() + scale_x_log10() +scale_y_log10(limits = c(0.01, 0.5), n.breaks = 8) 
#+ geom_vline(xintercept = 1e-5, linetype = "dashed") + geom_vline(xintercept = 1e-9, linetype= "dashed")#+ylim(0,0.3)
p2 = ggplot(dt,aes(x= FractionInfectedDaily, y = CapacityNeeded, colour = as.factor(TestPeriod))) + geom_line() + scale_x_log10() +scale_y_log10(limits = c(0.01, 1), n.breaks = 8) 
#+ geom_vline(xintercept = 1e-5, linetype = "dashed") + geom_vline(xintercept = 1e-9, linetype= "dashed")#+ylim(0,0.3)

print(p1)
print(p2)

```
Figure 3: Daily cost of strategy depending on disease prevalence. Split by isolation cost and testing cost. Fig b: capacity requirement with prevalence. todo: show Re for each strategy in legend?




todo: justify probability of export between regions
todo: justify detection size





https://stats.stackexchange.com/questions/94311/does-the-estimated-overdispersion-parameter-of-negative-binomial-depend-on-mean
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7338915/

rnbinom(1, size = 0.1, mu = 3 )


```{r}
sampleNumGenerations = function(Re, overdispersion=0.1, initialSize = 1){
  if(Re >= 1.0) return(NA)
  numInfected = initialSize
  
  count = 0
  totalInfected = 0
  while(count < 1e4 && numInfected > 0){
    totalInfected = totalInfected + numInfected
    count = count + 1
    numInfected = sum(rnbinom(numInfected, size = overdispersion, mu = Re ))
  }
  return(data.table(NumGenerations = count, TotalInfected = totalInfected ))
}


estimateOutbreakDuration = function(Re, initialSize, tau = 5, safetyMargin = 10){
  numGen = mean(sapply(1:10000, function(x) {dt = sampleNumGenerations(Re = Re, initialSize = initialSize); return(dt$NumGenerations)}))
  return(tau*numGen + safetyMargin)
}
#mean(sapply(1:10000, function(x) {dt = sampleNumGenerations(Re = 0.9, initialSize = 10); return(dt$TotalInfected)}))


steadyImportCost = function(importRate, popSize, Re, controlTime, dailyCost, infectionCost, detectionSize,  fracPopTarget, probExportPerInfected){
  
  numDailyImports = importRate*popSize
  
  if(is.na(controlTime)){
    return(dailyCost + infectionCost*importRate*1/(1-Re))
  }else{
    outbreakSize = detectionSize/(1-Re)
    outbreakCost = infectionCost*outbreakSize/popSize + fracPopTarget*controlTime*dailyCost
    
    if(probExportPerInfected*outbreakSize >= 1){
      
      return(dailyCost + infectionCost*importRate*1/(1-Re))
    }
    
    
    expectedNewOutbreaks = 1/(1 - probExportPerInfected*outbreakSize)
    

    
    
    return(numDailyImports*outbreakCost*expectedNewOutbreaks) # todo: consider recursive outbreak seeding
  }
}

popSize = 60e6
fracDt = fread("~/MassTesting/FracTransmissions.csv")
R0 = 2
incentiveCost = 5
logisticCost = 8
poolSize = 24
pcrCost = 48
isoCost = 5e3
testFraction = 0.95
isoFraction = 0.9
testPeriod = 2
testDelay = 1

dt = rbindlist(llply(10^seq(-11, -2, length.out = 200), function(importRate){
  
  transFraction = fracDt[TestPeriod == testPeriod & TestDelay == testDelay, FracAfterPositive]
  
  testFreq = 1/testPeriod
  
  Re = R0*(1 - testFraction*transFraction*isoFraction)
  
  if(Re > 1){
    cost = NA
    dailyTestCostPerPerson = NA
    dailyIsoCostPerPerson = NA
  }
  else{
      newInfectFrac = importRate*1/(1-Re) # fraction new infections (import and transmit per day) relative to total pop
  
    if(poolSize == 1){
      testCost = testFreq*testFraction*(incentiveCost + logisticCost + pcrCost)
    }else{
      testCost = testFreq*testFraction*(incentiveCost + logisticCost + 1/poolSize*pcrCost + poolSize*newInfectFrac*testPeriod*pcrCost)
    }
      
    gdpPerCapita = 70e3 # gdp per person in USA
    
    dailyTestCostPerPerson = testCost*365/gdpPerCapita
    dailyIsoCostPerPerson = newInfectFrac*testFraction*isoCost*365/gdpPerCapita
    
    cost = dailyTestCostPerPerson + dailyIsoCostPerPerson
  }
  

  return(data.table(Cost = cost, TestCost = dailyTestCostPerPerson, IsoCost = dailyIsoCostPerPerson, Re = Re, TestPeriod = testPeriod, ImportRate = importRate, TestDelay = testDelay, FractionInfectedDaily = importRate/(1-Re)))

} ))

dt[, Strategy := "Continuous Testing"]




dtTimeTarget = copy(dt)
dtTimeTarget[, Strategy := "Temporal Targeting"]
dtTimeTarget[, FracPopTarget := 1]
dtTimeTarget[, ProbExportPerInfected := 0]

dtSpatioTempTarget = copy(dt)
dtSpatioTempTarget[, Strategy := "Spatio-temporal Targeting"]
dtSpatioTempTarget[, FracPopTarget := 10e3/60e6]
dtSpatioTempTarget[, ProbExportPerInfected := 0.02]


dtTarget = rbind(dtSpatioTempTarget, dtTimeTarget)


detectionSize = 10
dtTarget[ , LocalOutbreakDuration := estimateOutbreakDuration(Re, detectionSize, 5, 10), by = Re]

dtTarget[, Cost := steadyImportCost(ImportRate, popSize, Re, LocalOutbreakDuration, TestCost, 5e3/70e3, detectionSize, FracPopTarget, ProbExportPerInfected), by = list(Re, LocalOutbreakDuration, TestCost, ImportRate, TestPeriod, TestDelay,FracPopTarget, ProbExportPerInfected )]

dt = rbind(dt, dtTarget, fill = TRUE)

dt[,TestPeriod := as.factor(TestPeriod)]
ggplot(dt[ TestPeriod ==2 & TestDelay == 1],aes(x= ImportRate, y = Cost, colour =Strategy)) + geom_line() + scale_x_log10() +scale_y_log10(limits = c(0.0002, 0.5), n.breaks = 8) + geom_vline(xintercept = 1e-5, linetype = "dashed") + geom_vline(xintercept = 1e-9, linetype= "dashed") #+ theme(legend.position = "bottom") #+ylim(0,0.3)



```
Figure 4: daily cost of differently targeted frequent testing strategies


Can be very cost effective for some diseases. How likely are these diseases?


Todo:
Fig 5: R0 vs mean incubation, with curves for diff testing params such that Re = 1. Dots for various diseases


Todo: describe usefulness even when Re from testing > 1


Todo: describe potential implementation in detail



\iffalse
#Notes

With a PCR cost of \$40 per test, and \$10 incentive, testing everyone every 2 days with a 1 day test delay costs about 12\% of GDP when importation rate below 1e-4. This cost is dominated by the cost of testing. Only when imporation rate exceeds 1e-4 (in this scenario) that cost of incentivizing and supporting isolation becomes significant. 



12\% GDP worthwhile for deadly pandemic, but quite dificult to mobilize, and harder to make the case for apparently less severe pandemics. Can we reduce this cost?


focus on convenience of intervention so cost of point of indifference drops from \$10 to \$5. 
use tests that are more comfortable so price point of indifference lower (saliva rather than nasal)


pooled testing used in many places [] during Covid, but there was limited time to develop optimized systems and the disease prevalence was quite high, so many tests had to be repeated [], increasing the complexity and reducing cost saving from pooled testing

however, in a well-controlled pandemic, the disease prevalence would be so low that the fraction of repeated tests would be insignificant

Testing cost = (1/20 pcrCost  + 20 FracPositive * pcrCost + transportationCost + incentiveCost) * numTests
along with additional 4 hour delay, and slightly reduced sensitivity []

fracPositive less than importRate*1/(1-Rt) = R0*1/(1 - testFraction*transFraction*isoFraction)




pooled testing to bring pcr cost down from \$40 to \$5 or less. 
with \$10 per test would be 2.4\% GDP - much closer to being an easy choice.


Todo:
Vary, R0, incubation period. budget? hold import rate fixed at 1e-5?
cost as function of prevalence when import rate fixed? point of insuficient capacity or budget?



For testing every 2 days, 
50 USD capital cost every 10 years per person. Or 5 USD per person per year to have sufficient pcr equipment ready (but not necessarily space or trained technicians)
500k USD yearly salary for lab staff in India? maybe 5x more expensive in US or UK? 2.5M USD per year to have full staff on standby. 
=  625USD per person per year (way too high)

87% of 7.5 USD  = 6.5USD cost is laboratory consumables (rRT-PCR reagents and kits, plasticware and glass wares)


# Extensions
partial contact tracing:

simple version is assume independence, but people who transmit likely transmit early, so have short incub? correlated?

monte carlo branching model. sample testing offset from infection, sample incubation, sample test results, sample transmissions, sample incubation of infectees, sample testing offset, sample infectee test results. sum expected second degree transmissions.

Track number of second degree infections. Average should be Re^2 ?

\fi

